The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-01-19 paper.tex
restructure

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
%\usepackage[T1]{fontenc}
\usepackage{todonotes}
\usepackage{cleveref}
\usepackage{csquotes}
\usepackage{aurl}
\usepackage{booktabs}
\usepackage{tabulary}
\usepackage{siunitx}
%\usepackage{dcolumn}
%\usepackage{placeins} %provides \FloatBarrier

@@

%\paragraph{Theoretical Background}
Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.
Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
The success of using ontologies to annotate biological data can also be confirmed by multiple examples~\citep{biomedicalontologies}.
It was therefore an easy decision for us to use an ontology for modeling the anthropological domain.

%\paragraph{Tasks}
The main tasks to be supported by the intended ontology are the annotation of anthropological models, the specification of spatial relations between anatomical entities and the definition of phenotyping functions, see \cref{sec:methodology}.
However, the ontology must remain relatively simple and compact to allow easy handling by domain experts and efficient integration into software.
Existing anatomical ontologies often lack standardization in the naming and definition of concepts.
Furthermore, the hierarchical structure is often too complex/general to be efficiently integrated into software for realizing specific anthropological use cases, see \cref{sec:limitations}.
Since we could not find any ontologies that fully met our requirements and the integration/adaptation of selected ontology parts would be much more complex, we decided to develop a new ontology and then map it to the most prominent one.

%\paragraph{Main contributions}
% rest of old text
%ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
%
%new text:
Our main contributions are the following:
ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes. ANNOdc was embedded in the General Formal Ontology (GFO), i.e. the ANNOdc classes are subclasses of GFO classes.
%
ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between Mentale dexterum and Mentale sinistrum, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
%
In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.

\begin{table}[h]
\centering
\caption{Structure, properties and publication of ANNO.}
\label{tab:publication}
\begin{tabulary}{\textwidth}{ll}
\toprule
%Property			&Value\\
%\midrule
Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
Namespace			&\url{https://annosaxfdm.de/ontology/}\\
Browser			&RickView~\citep{rickview} at \url{https://annosaxfdm.de/ontology/}\\
Repository		&\url{https://github.com/annosaxfdm/ontology}\\
Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
License				&\href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International}\\
DOI (all versions)		&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
\bottomrule
\end{tabulary}
\end{table}

@@

\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}
The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top-level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.

@@

Uberon, the Uber Anatomy Ontology~\citep{uberon}...
BSPO~\citep{bspo}...
NCIT~\citep{ncit}...
The Anatomical Entity Ontology (AEO)~\citep{aeo}... (for animals)
Read Codes Version 3 (RCD)~\citep{rcd}...

@@

%\subsection{Current Research Gaps}
There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.

The primary issue lies in the absence of standardization.
While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in actual practice~\citep{frequencyta,doestamatter,athighlights}.
Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
Yet, knowledge of Latin is generally declining.
With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
Moreover, in textual descriptions Latin and the native language are commonly used interchangeably for stylistic purposes.
For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.

@@

\section{ANthropological Notation Ontology (ANNO)}\label{sec:anno}
\subsection{Methodology}\label{sec:methodology}
\subsection{Description and foundation of ANNOdc}\label{sec:annodc}
%Terminology for descriptive anatomy
%Anatomical entities
%Phenotypes and anatomical Properties

\subsection{Development of ANNOds}\label{sec:annods}

\section{Ontological Architecture}\label{sec:architecture}
\section{Proof of concept: Integration into \aw{}}\label{sec:integration}

The ontological architecture of ANNO consists of three interrelated layers representing by a top-level ontology, a domain-core ontology and a domain-specific ontology:

\begin{enumerate}
\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},

\item ANNOdc (\cref{sec:core}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
ANNOdc is embedded in the GFO, i.e. the ANNOdc classes are subclasses of GFO classes.

\item ANNOds (\cref{sec:domain}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
\end{enumerate}

@@

\section{Methodology for developing ANNO}
For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology \citep{baumann2014, herre2010}.
The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain \citep{baumann2014, herre2010}.
When knowledge is systematised, a set of categories is considered \emph{primitive} or \emph{basic}.
Such categories are not explicitly defined, but implicitly described by axioms \citep{Hilbert1918}.
An example of a primitive category is \enquote{part}.
New notions can be introduced by explicit definitions based on the primitive or already defined categories \citep{herre2010}.

@@

\subsection{Domain specification}
As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
Together with ontologists, relevant use cases and competence questions\footnotemark{} \citep{XD2016, MOMo2023}, as well as views and classification principles of the objects in the anthropology domain \citep{herre2010} were discussed.
\footnotetext{queries that the ontology must be able to answer}
%
The following main use cases and competence questions (sub-items, selected examples) were defined:
\begin{enumerate}
\item Annotation of anthropological models (representations of anatomical entities) using a controlled vocabulary
\begin{enumerate}
\item Query all bone types
\item Query all parts (types) of a specific bone (type)
\item Query all parts (types) of a specific bone compound (type)
\item Query all tooth types
\item Query all parts (types) of a specific tooth (type)
\item $\ldots$
\end{enumerate}
\item Specification of spatial relations between anatomical entities
\begin{enumerate}
\item Query all defined spatial relations
\item What is the relative anatomical location of an anatomical entity in relation to another anatomical entity?
\item On which anatomical structure does a defined measurement point lie?
\item Between which measurement points does a defined anatomical line lie?
\item $\ldots$
\end{enumerate}
\item Specification of phenotyping functions
\begin{enumerate}
\item Query all phenotype specifications
\item Query the phenotyping function for determining a particular phenotype
\item Which parameters/variables (e.g., measurement distances/angles/points) are required to determine a particular phenotype?
\item $\ldots$
\end{enumerate}
\end{enumerate}
Relevant domain objects were identified on the basis of the defined use cases.
These include bones, teeth, phenotypes, as well as spatial anatomical entities.
The objects are classified according to their type and part-whole relationship.
The main requirement for the ANNO core ontology to cover the intended use cases is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method~\citep{threeontologymethod}. 
Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.\todo{Absatz doppelt (Ende \cref{sec:limitations})}
We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see \cref{sec:relatedwork}).

\subsection{Conceptualisation}

During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
The concepts were created by generalising and classifying the domain objects.
To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
Therefore, the corresponding concepts \enquote{Bone}, \enquote{Bone part} and \enquote{Bone compound} were introduced.
Further concepts were defined to represent spatial relations and phenotypes.
Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:annodc}.

\subsection{Axiomatization}

For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space} \citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
For example, GFO specifies that material objects occupy space regions \citep{Loebe2021}.
We integrated this pattern in ANNOdc by deriving the class anno:AnatomicalStructure from gfo:MaterialObject and anno:AnatomicalSpace from gfo:SpaceRegion.
In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region} \citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
According to the GFO relator \citep{Loebe2018} pattern, we defined the ANNO class \enquote{Relative anatomical location} including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).

@@

The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT and Pellet reasoners~\citep{hermit,pellet} and SHACL shapes (see \cref{sec:domain}).
In both cases, the consistency of the ontology was proven.
As a proof of concept, we integrated the ontology into the AW3D software~\citep{aw3d} and were able to successfully realise all three intended use cases.
Thereby, most of the core categories (such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}) and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).

@@

\section{Description and Foundation of ANNOdc}\label{sec:core}
The core ontology development occurs in close collaboration between ontologists and anthropologists.
The objective is twofold: to adequately represent the most basic anatomical and anthropological entities and to keep the ontology relatively simple and compact.
This approach ensures that domain experts can easily handle the ontology and efficiently integrate it into the \aw{} software, see \cref{sec:aw}.

@@

\subsection{Terminology for descriptive anatomy}

@@

\subsection{Anatomical entity and structure}
An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.

An anatomical structure refers to any anatomical division or material anatomical entity.
ANNO describes three kinds of anatomical structures: tooth structures (teeth and teeth parts), bone structures (single bones, bone parts and bone compounds), as well as the complete skeleton.
The GFO defines a material object \aurl{gfo}{MaterialObject} as a solid concrete entity that belongs to the material region of the world, has mass, consists of matter and occupies space~\citep{gfospace}.
Accordingly, we define anno:AnatomicalStructure as a gfo:MaterialObject in the anthropological context.
The closest equivalent in the FMA is the \emph{Material anatomical entity} \aurl{fma}{67165}, the subclass of \emph{Physical anatomical entity} \aurl{fma}{61775} that has mass.

\subsection{Skeleton}
The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human's bones and teeth.
However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g. as specific aspects), to be integrated and referenced separately in ANNO.
The anthropologically meaningful distinction between \enquote{Skeleton} and \enquote{Systema skeletale} can be maintained, while still referencing both within the TA concept.

\subsection{Tooth structure}
Tooth structures comprise teeth and their parts.
A human tooth is an individual unit of the human dentition (synonymous for the teeth as a whole).
Teeth are part of the skeleton, yet they are characterized by their own distinctive tissue and thus treated separately from bones.
Teeth, such as the Dens caninus (canine tooth) are located on bone structures.
The FMA has the equivalent class \enquote{Tooth} \aurl{fma}{12516}.
A tooth part is any portion of a tooth.
They are not included in the initial scope of the ontology, but they are one of the aspects by which it can be expanded.

\subsection{Bone structure}
Bone structures comprise all possible parts of the human skeleton, which are individual bones, bone parts, bone compounds as well as the complete skeleton excluding the teeth.
The FMA does not have a common superclass for those parts but instead mounts them in different parts of its hierarchy.

@@

\subsection{Anatomical property}
An anatomical property is a characteristic (e.g., shape) of an anatomical structure.
Anatomical properties are considered as attributes or qualities in GFO~\citep{gfoarchitecture}.
These are dependent individuals that characterise other individuals (in our case, anatomical structures).

>>>>

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
%\usepackage[T1]{fontenc}
\usepackage{todonotes}
\usepackage{cleveref}
\usepackage{csquotes}
\usepackage{aurl}
\usepackage{booktabs}
\usepackage{tabulary}
\usepackage{siunitx}
\usepackage{tablefootnote}
%\usepackage{dcolumn}
%\usepackage{placeins} %provides \FloatBarrier

@@

%\paragraph{Theoretical Background}
\todo{der Absatzübergang ist etwas holprig}
Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.
Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
The success of using ontologies to annotate biological data can also be confirmed by multiple examples~\citep{biomedicalontologies}.
It was therefore an easy decision for us to use an ontology for modeling the anthropological domain.

%\paragraph{Tasks}
The main tasks to be supported by the intended ontology are the annotation of anthropological models, the specification of spatial relations between anatomical entities and the definition of phenotyping functions, see \cref{sec:methodology}.
However, the ontology must remain relatively simple and compact to allow easy handling by domain experts and efficient integration into software.
Existing anatomical ontologies often lack standardization in the naming and definition of concepts.
Furthermore, the hierarchical structure is often too complex or general to be efficiently integrated into software for realizing specific anthropological use cases, see \cref{sec:limitations}.
Since we could not find any ontologies that fully met our requirements and the integration or adaptation of selected ontology parts would be much more complex, we decided to develop a new ontology and then map it to the most prominent one.

\begin{table}[h]
\centering
\caption{Structure, properties and publication of ANNO.}
\label{tab:publication}
\begin{tabular}{ll}
\toprule
%Property			&Value\\
%\midrule
Modules				&ANNOdc (domain-core) , ANNOds (domain-specific)\\
Top-Level Ontology	&General Formal Ontology~\citep{gfo}\\
Namespace			&\url{https://annosaxfdm.de/ontology/}\\
%Browser			&RickView~\citep{rickview} at \url{https://annosaxfdm.de/ontology/}\\
Browser				&RickView\tablefootnote{\url{https://github.com/KonradHoeffner/rickview}} at \url{https://annosaxfdm.de/ontology/}\\
Repository			&\url{https://github.com/annosaxfdm/ontology}\\
Terminology Server	&\url{https://ols.imise.uni-leipzig.de/ontologies/anno}\\
License				&\href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution 4.0 International}\\
DOI (all versions)	&\href{https://doi.org/10.5281/zenodo.8301559}{10.5281/zenodo.8301559}\\
\bottomrule
\end{tabular}
\end{table}

%\paragraph{Main contributions}
% rest of old text
%ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures.
%
%new text:
Our main contributions are the following:
ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category Phenotype to model the rules for determining human phenotypes. ANNOdc was embedded in the General Formal Ontology (GFO), i.e. the ANNOdc classes are subclasses of GFO classes.
%
ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as mandible, Mental protuberance or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
%
In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains, and published over multiple channels, see~\cref{tab:publication}.

@@

\subsection{Foundational Model of Anatomy (FMA)}\label{sec:fma}\todo{an anderer Stelle haben wir für viel größere Textmengen keine subsection und hier für ganz wenig Text eine}
The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top-level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.

@@

\todo{Mittweida, bitte sobald wie möglich hier eintragen, es ist nicht mehr viel Zeit}
Uberon, the Uber Anatomy Ontology~\citep{uberon}...
BSPO~\citep{bspo}...
NCIT~\citep{ncit}...
The Anatomical Entity Ontology (AEO)~\citep{aeo}... (for animals)
Read Codes Version 3 (RCD)~\citep{rcd}...

@@

%\subsection{Current Research Gaps}
There are several issues and research gaps when it comes to naming and defining anatomical concepts and instances that neither current ontologies nor reference literature---general or subject-specific---can address adequately.
%
The primary issue lies in the absence of standardization.
While both the TA and FMA have been proposed as a means of standardizing terminology, they haven't garnered sufficient acceptance to facilitate the adoption of a unified and consistent body of work to establish standardization in actual practice~\citep{frequencyta,doestamatter,athighlights}.
Instead, the usage of terminology is as fluent as any other language, depending on its socio-cultural environment, such as schools or language areas~\citep{doestamatter,atthennow,atinfo,frequencyta,atcompare}.
Thus, the usage of English designations is preferred in the Anglo-American sphere while Latin terms are commonly used in German-speaking areas~\citep{anatomycontribution,anatomylexicon,reforminganatomical}.
As a consequence there are at times various translations for the same anatomical terms~\citep{naminggame}.
Yet, knowledge of Latin is generally declining.
With the terms hence becoming more abstract to the people using them, there is also a common prevalence for spelling divergences or grammar mistakes~\citep{ta17,anatomylexicon,athighlights,diphthongs}.
Moreover, in textual descriptions Latin and the native language are commonly used interchangeably for stylistic purposes.
For instance, \cite{anatomylexicon} uses the German terms of the bones in the explanation of sutures.
Meanwhile, the existing terminology is not flexible enough for language-like usage as, for example, there is an inconsistent use of singular and plural forms or gaps in lateralized landmarks.

@@

\section{ANthropological Notation Ontology (ANNO)}\label{sec:anno}

\subsection{Methodology}\label{sec:methodology}
The ontological architecture of ANNO consists of three interrelated layers representing by a top-level ontology, a domain-core ontology and a domain-specific ontology:

\begin{enumerate}
\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:annodc},

\item ANNOdc (\cref{sec:annodc}, \cref{fig:annodc}) is a domain-core (dc) ontology that defines the core entities of the domain.
This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
ANNOdc is embedded in the GFO, i.e. the ANNOdc classes are subclasses of GFO classes.

\item ANNOds (\cref{sec:annods}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
It is also used for modeling their properties and relations, such as the distance between \emph{Mentale dexterum} and \emph{Mentale sinistrum}, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
\end{enumerate}

@@

%\section{Methodology for developing ANNO}
For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology \citep{baumann2014, herre2010}.
The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain \citep{baumann2014, herre2010}.
When knowledge is systematised, a set of categories is considered \emph{primitive} or \emph{basic}.
Such categories are not explicitly defined, but implicitly described by axioms \citep{Hilbert1918}.
An example of a primitive category is \enquote{part}.
New notions can be introduced by explicit definitions based on the primitive or already defined categories \citep{herre2010}.

@@

\paragraph{Domain specification}
As part of the domain specification, the anthropology experts conducted an extensive review of existing literature and ontologies, analysing and classifying the relevant information.
Together with ontologists, relevant use cases and competence questions\footnotemark{} \citep{XD2016, MOMo2023}, as well as views and classification principles of the objects in the anthropology domain \citep{herre2010} were discussed.
\footnotetext{queries that the ontology must be able to answer}
%
The following main use cases and competence questions (sub-items, selected examples) were defined:
\begin{enumerate}
\item Annotation of anthropological models (representations of anatomical entities) using a controlled vocabulary
\begin{enumerate}
\item Query all bone types
\item Query all parts (types) of a specific bone (type)
\item Query all parts (types) of a specific bone compound (type)
\item Query all tooth types
\item Query all parts (types) of a specific tooth (type)
\item $\ldots$
\end{enumerate}
\item Specification of spatial relations between anatomical entities
\begin{enumerate}
\item Query all defined spatial relations
\item What is the relative anatomical location of an anatomical entity in relation to another anatomical entity?
\item On which anatomical structure does a defined measurement point lie?
\item Between which measurement points does a defined anatomical line lie?
\item $\ldots$
\end{enumerate}
\item Specification of phenotyping functions
\begin{enumerate}
\item Query all phenotype specifications
\item Query the phenotyping function for determining a particular phenotype
\item Which parameters/variables (e.g., measurement distances/angles/points) are required to determine a particular phenotype?
\item $\ldots$
\end{enumerate}
\end{enumerate}
Relevant domain objects were identified on the basis of the defined use cases.
These include bones, teeth, phenotypes, as well as spatial anatomical entities.
The objects are classified according to their type and part-whole relationship.
The main requirement for the ANNO core ontology to cover the intended use cases is that it adequately represents the basic anatomical and anthropological entities, the spatial relations between them and the phenotyping functions, while remaining relatively simple and compact.
This approach allows on the one hand an easy manageability of the ontology for domain experts and on the other hand its efficient integration into software (e.g., AW3D) according to the three-ontology method~\citep{threeontologymethod}. 
Following this method, the software only needs to implement access to the core categories (of the core or task ontology), while their subcategories (from the domain-specific ontology) are processed dynamically.
This means, among other things, that all relevant anatomical entities (bones, teeth, their parts and composite structures) must each be located in one subtree (i.e., must have a common superclass) rather than spread across the entire ontology.
Furthermore, clear relations must exist between the core entities so that the dependencies between them can be efficiently resolved by the software.\todo{Absatz doppelt (Ende \cref{sec:limitations})}
We decided to develop a new ontology because none of the existing ontologies could cover all defined requirements and integrating (extending and adapting) all the necessary parts from the other ontologies would have been much more complex (see \cref{sec:relatedwork}).

\paragraph{Conceptualisation}
During the conceptualisation phase \citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
The concepts were created by generalising and classifying the domain objects.
To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
Therefore, the corresponding concepts \enquote{Bone}, \enquote{Bone part} and \enquote{Bone compound} were introduced.
Further concepts were defined to represent spatial relations and phenotypes.
Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:annodc}.

\paragraph{Axiomatization}
For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns \citep{ODP2005, XD2016, MOMo2023}).
We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space} \citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
For example, GFO specifies that material objects occupy space regions \citep{Loebe2021}.
We integrated this pattern in ANNOdc by deriving the class anno:AnatomicalStructure from gfo:MaterialObject and anno:AnatomicalSpace from gfo:SpaceRegion.
In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region} \citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
According to the GFO relator \citep{Loebe2018} pattern, we defined the ANNO class \enquote{Relative anatomical location} including the two roles: target (locationOf) and reference (relativeTo) anatomical entity (chapter $\ldots$).

@@

The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT~\citep{hermit} and Pellet~\citep{pellet} reasoners and SHACL shapes, see \cref{sec:annods}.
In both cases, the consistency of the ontology was proven.
As a proof of concept, we integrated the ontology into the AW3D software~\citep{aw3d} and were able to successfully realise all three intended use cases.
Thereby, most of the core categories,
such as \enquote{Bone}, \enquote{Bone part}, \enquote{Bone compound}, \enquote{Tooth}, \enquote{Tooth part}, \enquote{Phenotype}, \enquote{Relative anatomical location}, \enquote{Anatomical line} and \enquote{Anatomical point}),
and relation types (such as hasPart, boundaryOf, locationOf and derivedFrom) of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).

@@

\subsection{Description and foundation of ANNOdc}\label{sec:annodc}
The core ontology development occurs in close collaboration between ontologists and anthropologists.
The objective is twofold: to adequately represent the most basic anatomical and anthropological entities and to keep the ontology relatively simple and compact.
This approach ensures that domain experts can easily handle the ontology and efficiently integrate it into the \aw{} software, see \cref{sec:aw}.

@@

%\paragraph{Terminology for descriptive anatomy}

@@

%\paragraph{Anatomical entities}
An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.
An anatomical structure refers to any anatomical division or material anatomical entity.
ANNO describes three kinds of anatomical structures: tooth structures (teeth and teeth parts), bone structures (single bones, bone parts and bone compounds), as well as the complete skeleton.
The GFO defines a material object \aurl{gfo}{MaterialObject} as a solid concrete entity that belongs to the material region of the world, has mass, consists of matter and occupies space~\citep{gfospace}.
Accordingly, we define anno:AnatomicalStructure as a gfo:MaterialObject in the anthropological context.
The closest equivalent in the FMA is the \emph{Material anatomical entity} \aurl{fma}{67165}, the subclass of \emph{Physical anatomical entity} \aurl{fma}{61775} that has mass.

%\paragraph{Skeleton}
The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
ANNO does not adopt this equivalence, and \enquote{Skeleton } here refers to the entirety of a human's bones and teeth.
However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g. as specific aspects), to be integrated and referenced separately in ANNO.
The anthropologically meaningful distinction between \enquote{Skeleton} and \enquote{Systema skeletale} can be maintained, while still referencing both within the TA concept.

%\paragraph{Tooth structure}
Tooth structures comprise teeth and their parts.
A human tooth is an individual unit of the human dentition (synonymous for the teeth as a whole).
Teeth are part of the skeleton, yet they are characterized by their own distinctive tissue and thus treated separately from bones.
Teeth, such as the Dens caninus (canine tooth) are located on bone structures.
The FMA has the equivalent class \enquote{Tooth} \aurl{fma}{12516}.
A tooth part is any portion of a tooth.
They are not included in the initial scope of the ontology, but they are one of the aspects by which it can be expanded.

%\paragraph{Bone structure}
Bone structures comprise all possible parts of the human skeleton, which are individual bones, bone parts, bone compounds as well as the complete skeleton excluding the teeth.
The FMA does not have a common superclass for those parts but instead mounts them in different parts of its hierarchy.

@@

%\paragraph{Anatomical property}
An anatomical property is a characteristic (e.g., shape) of an anatomical structure.
Anatomical properties are considered as attributes or qualities in GFO~\citep{gfoarchitecture}.
These are dependent individuals that characterise other individuals (in our case, anatomical structures).
